---
title: "Severe weather events analysis"
output: html_document
---

## Synopsis
This data analysis concentrates on answering two questions: 1) Which types of weather events have the most harmful effect on the population health? 2) Which types of the weather events have the greatest economical consequences? Analysis is based on the data gathered by U.S. National Oceanic and Atmospheric Administration. 

## Data Processing
Source data needs to be downloaded from the following link: [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

File needs to be extracted using bzip2 utility

Content can now be read into R
```{r echo=TRUE, cache=TRUE}
setwd("~/R/represearch")
stormdata<-read.table(file="repdata-data-StormData.csv", sep=",", header=TRUE)
#head(stormdata)
#str(stormdata)
dim(stormdata)
#levels(stormdata$PROPDMGEXP)
stormdata<-transform(stormdata,EVTYPE=as.factor(toupper(EVTYPE))) 
length(levels(stormdata$EVTYPE))
```

### 1. Extract only data relevant for health issues
```{r cache=TRUE}
stormhealth<-  subset ( stormdata, FATALITIES > 0 | INJURIES > 0 , select = c(EVTYPE,FATALITIES,INJURIES) )
dim(stormhealth)
stormhealth$EVTYPE = as.factor (as.character (stormhealth$EVTYPE))
#levels(stormhealth$EVTYPE)
#str(stormhealth)
```
From the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) 2.1.1 Storm Data Event Table: extract the official event names and include into the analysis manually.
```{r}
evt_name<-c("Astronomical Low Tide","Avalanche","Blizzard","Coastal Flood","Cold/Wind Chill","Debris Flow","Dense Fog","Dense Smoke","Drought","Dust Devil","Dust Storm","Excessive Heat","Extreme Cold/Wind Chill","Flash Flood","Flood","Frost/Freeze","Funnel Cloud","Freezing Fog","Hail","Heat","Heavy Rain","Heavy Snow","High Surf","High Wind","Hurricane (Typhoon)","Ice Storm","Lake-Effect Snow","Lakeshore Flood","Lightning","Marine Hail","Marine High Wind","Marine Strong Wind","Marine Thunderstorm Wind","Rip Current","Seiche","Sleet","Storm Surge/Tide","Strong Wind","Thunderstorm Wind","Tornado","Tropical Depression","Tropical Storm","Tsunami","Volcanic Ash","Waterspout","Wildfire","Winter Storm","Winter Weather")

evt_name <- toupper(evt_name)
#length(evt_name)
```
Find how many event names from the official manual match the event names in the csv file
```{r}
matches <-  (data.frame ( EVTYPE=stormhealth$EVTYPE, EVT_NAME=evt_name[match(stormhealth$EVTYPE, evt_name)] ))
summary(matches)
mean( is.na(matches$EVT_NAME ))
#str(matches)
```
~4700 or 20% events are unmatched. This is quite a high number. Need to find out which ones didn't match. 
```{r}
matches$count <- 1
matches$EVT_NAME<-as.character(matches$EVT_NAME)
matches$EVT_NAME[is.na(matches$EVT_NAME)] <- c("none") 
x<-aggregate( count ~ EVTYPE + EVT_NAME , data=matches, FUN=sum )
#summary(x)
#str(x)
x<-x[order(x$count),]
```
Top unmatched event types are:
```{r}
tail( x[x$EVT_NAME=='none',])
```
Will fix the top unmatched Event types
```{r}
stormhealth$EVTYPE[stormhealth$EVTYPE=="TSTM WIND"]<-"THUNDERSTORM WIND"
stormhealth$EVTYPE[stormhealth$EVTYPE=="THUNDERSTORM WINDS"]<-"THUNDERSTORM WIND"
stormhealth$EVTYPE[stormhealth$EVTYPE=="RIP CURRENTS"]<-"RIP CURRENT"
stormhealth$EVTYPE[stormhealth$EVTYPE=="WILD/FOREST FIRE"]<-"WILDFIRE"
stormhealth$EVTYPE[stormhealth$EVTYPE=="EXTREME COLD"]<-"EXTREME COLD/WIND CHILL"
stormhealth$EVTYPE[stormhealth$EVTYPE=="HIGH WINDS"]<-"HIGH WIND"
matches <-  (data.frame ( EVTYPE=stormhealth$EVTYPE, EVT_NAME=evt_name[match(stormhealth$EVTYPE, evt_name)] ))
```
Percentage of unmatched values is now:
```{r}
mean( is.na(matches$EVT_NAME ))
```
This is acceptible for further research

Now the data will be aggreated and top 10 event types with respect to human health displayed.

```{r}
agg_storm_h<-aggregate ( FATALITIES+INJURIES~EVTYPE, data=stormhealth, FUN=sum )
agg_storm_h<-agg_storm_h[ order( agg_storm_h$"FATALITIES + INJURIES" , decreasing = TRUE ),]
head(agg_storm_h, n=10)
```

### 2. Extract data relavant for economical consequences

Extract econimicaly relevant data
```{r cache=TRUE}
stormeconomic <-  subset ( stormdata, PROPDMG > 0 | CROPDMG > 0 , select = c(EVTYPE,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP) )
dim(stormeconomic)
stormeconomic$EVTYPE = as.factor (as.character (stormeconomic$EVTYPE))
```

Find the top unmatched events
```{r}
matches <-  (data.frame ( EVTYPE=stormeconomic$EVTYPE, EVT_NAME=evt_name[match(stormeconomic$EVTYPE, evt_name)] ))
summary(matches)
mean( is.na(matches$EVT_NAME ))
```
Approx 32% didn't match. Let's find out which ones are in top
```{r}
matches$count <- 1
matches$EVT_NAME<-as.character(matches$EVT_NAME)
matches$EVT_NAME[is.na(matches$EVT_NAME)] <- c("none") 
x<-aggregate( count ~ EVTYPE + EVT_NAME , data=matches, FUN=sum )
#summary(x)
#str(x)
x<-x[order(x$count),]
tail( x[x$EVT_NAME=='none',])
```
Will fix top ones
```{r}
stormeconomic$EVTYPE[stormeconomic$EVTYPE=="TSTM WIND"]<-"THUNDERSTORM WIND"
stormeconomic$EVTYPE[stormeconomic$EVTYPE=="THUNDERSTORM WINDS"]<-"THUNDERSTORM WIND"
stormeconomic$EVTYPE[stormeconomic$EVTYPE=="RIP CURRENTS"]<-"RIP CURRENT"
stormeconomic$EVTYPE[stormeconomic$EVTYPE=="WILD/FOREST FIRE"]<-"WILDFIRE"
stormeconomic$EVTYPE[stormeconomic$EVTYPE=="EXTREME COLD"]<-"EXTREME COLD/WIND CHILL"
stormeconomic$EVTYPE[stormeconomic$EVTYPE=="HIGH WINDS"]<-"HIGH WIND"
stormeconomic$EVTYPE[stormeconomic$EVTYPE=="TSTM WIND/HAIL"]<-"THUNDERSTORM WIND"
matches <-  (data.frame ( EVTYPE=stormeconomic$EVTYPE, EVT_NAME=evt_name[match(stormeconomic$EVTYPE, evt_name)] ))
summary(matches)
mean( is.na(matches$EVT_NAME ))
```

Unmathed events are down to ~2%. This is acceptable for further analysis.

Aggregate data to get the top 10 event types with highest econimical impact
```{r}
stormeconomic$totalpropdmg <- ifelse ( stormeconomic$PROPDMGEXP=='B', stormeconomic$PROPDMG * 10^9,
                                ifelse ( stormeconomic$PROPDMGEXP=='M', stormeconomic$PROPDMG * 10^6,
                                  ifelse ( stormeconomic$PROPDMGEXP=='K', stormeconomic$PROPDMG * 10^3,
                                    ifelse ( stormeconomic$PROPDMGEXP=='H', stormeconomic$PROPDMG * 10^2, 0 
                                    )
                                  )
                                )
                              )
stormeconomic$totalcropdmg <- ifelse ( stormeconomic$CROPDMGEXP=='B', stormeconomic$CROPDMG * 10^9,
                                ifelse ( stormeconomic$CROPDMGEXP=='M', stormeconomic$CROPDMG * 10^6,
                                  ifelse ( stormeconomic$CROPDMGEXP=='K', stormeconomic$CROPDMG * 10^3,
                                    ifelse ( stormeconomic$CROPDMGEXP=='H', stormeconomic$CROPDMG * 10^2, 0 
                                    )
                                  )
                                )
                              )
                                   
agg_econ <- aggregate ( (totalcropdmg + totalpropdmg)/10^6 ~ EVTYPE, data = stormeconomic, FUN=sum ) 
colnames( agg_econ)[2] <- "total_mln"
str(agg_econ)
agg_econ <- agg_econ[order(agg_econ$"total_mln", decreasing=TRUE),]
head(agg_econ,n=10)
```

## Results 
### 1. Top 10 events with respect to the population healt are:
```{r}
par(mai=c(1,3,1,1))
barplot ( head ( agg_storm_h$"FATALITIES + INJURIES", n=10) , main="top 10 health affecting events", names.arg=head(agg_storm_h$EVTYPE,n=10), xlab="Total Fatalities+Injuries" , horiz=TRUE , las=1)
```

### 2. Top 10 events with respect to economical impact are:
```{r}
par(mai=c(1,3,1,1))
barplot( head ( agg_econ$"total_mln", n=10), names.arg=head (agg_econ$EVTYPE,n=10),  main="Top 10 events by econimic impact" , las=1, horiz=TRUE, xlab='Total Million Dollar')
```






